
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading Libraries

```{r}
packages <- c("tidyverse", "REdaS", "psych", "FactoMineR", "factoextra", "writexl")
#lapply(packages, library, character.only = TRUE)
```

```{r, include=FALSE}
lapply(packages, library, character.only = TRUE)
```

### Loading and Preparing Data

```{r}
eyes <- read.table("F:/open_git/bee_eyes/data/eyes_df.txt", header = TRUE, sep = "\t")
```

#### Converting the FamHab column to a factor
```{r}
eyes$FamHab <- as.factor(eyes$FamHab)
```

#### Data summary
```{r}
summary(eyes)
```
### Principal Component Analysis (PCA)

##### Bartlett's test to check the adequacy of factor analysis
```{r}
cortest.bartlett(eyes[, 3:8])
```

### Calculating Principal Components

#### PCA without Rotation
```{r}
components_without_rotation <- principal(eyes[, 3:8], nfactors=2, rotate="none", scores=TRUE)
```
#### Visualizing the components
```{r}
components_without_rotation
```
### Biplot (No Rotation)
```{r}
biplot(components_without_rotation)
```

```{r}
components_without_rotation_1 <- PCA(eyes[, 3:8], scale.unit = TRUE, graph = TRUE, ncp=2)
```


```{r}
summary(components_without_rotation_1)
```
### PCA Metrics

#### eingvalues
```{r}
components_without_rotation$values
```
#### Variability Explained
```{r}
PC1 <- components_without_rotation$Vaccounted[2, 1]
PC2 <- components_without_rotation$Vaccounted[2, 2]
PC1; PC2

```
#### Factor Loadings and Scores
```{r}
components_without_rotation$weights
components_without_rotation$loadings
```
### Factorial Scores and Ranking

#### Calculating Factor Scores
```{r}
eyes_without_rotation <- eyes
eyes_without_rotation$scores_PCA1 <- components_without_rotation$scores[, 1]
eyes_without_rotation$scores_PCA2 <- components_without_rotation$scores[, 2]
```
#### Creating a Ranking Based on PCA Scores
```{r}
eyes_without_rotation <- eyes_without_rotation %>% 
  mutate(ranking = (PC1 * scores_PCA1) + (PC2 * scores_PCA2)) %>% 
  arrange(desc(ranking))

head(eyes_without_rotation)
```
#### Exporting the Ranking to Exce
```{r}
write_xlsx(eyes_without_rotation, "F:/open_git/bee_eyes/results/ranking_pca.xlsx")
```
### Custom Biplot Visualization
```{r}
biplot(components_without_rotation, pch = c(21, 21, 21, 21, 21), 
       col = c("orange", "purple", "brown3", "darkseagreen3", "dodgerblue3"), 
       group = eyes$FamHab, ylim.s = c(-4, 4), xlim.s = c(-4, 4))

legend("bottomleft", fill = c("orange", "purple", "brown3", "darkseagreen3", "dodgerblue3"), 
       horiz = FALSE, c("ColDiu", "ColFac", "ColCre", "HalCre", "HalDiu"), cex = 1, bty = "n")
```
