---
output:
  pdf_document: default
  html_document: default
---
# Clustering Analysis of Bee Eye Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading Libraries

```{r}
packages <- c("tidyverse", "cluster", "dendextend", "factoextra", "gridExtra", "writexl")
#lapply(packages, library, character.only = TRUE)
```

```{r, include=FALSE}
lapply(packages, library, character.only = TRUE)
```
### Function to perform hierarchical clustering

```{r}
perform_hierarchical_clustering <- function(data, method = "average", k = 3) {
  distance <- dist(data, method = "euclidean")
  cluster <- hclust(distance, method = method)
  plot(cluster, cex = 0.6, hang = -1)
  rect.hclust(cluster, k = k)  # Highlight groups
  return(cluster)
}
```

### Function to generate and plot clusters for K-means

```{r}
generate_kmeans_clusters <- function(data, centers, title) {
  kmeans_result <- kmeans(data, centers = centers)
  plot <- fviz_cluster(kmeans_result, geom = "point", data = data) + ggtitle(title)
  return(list(model = kmeans_result, plot = plot))
}
```

### Function to visualize elbow and silhouette methods

```{r}
evaluate_clusters <- function(data, clustering_function) {
  # Elbow method
  elbow_plot <- fviz_nbclust(data, clustering_function, method = "wss")
  print(elbow_plot)  # Exibe o gráfico do Elbow
  
  # Silhouette method
  silhouette_plot <- fviz_nbclust(data, clustering_function, method = "silhouette")
  print(silhouette_plot)  # Exibe o gráfico do Silhouette
}
```
### Function to calculate descriptive statistics for clusters

```{r}
calculate_means_by_group <- function(data, cluster_column) {
  summary <- data %>%
    group_by(!!sym(cluster_column)) %>%
    summarise(n = n(),
              ff = mean(ff),
              df = mean(df),
              vf = mean(vf),
              co = mean(co),
              nf = mean(nf),
              nfa = mean(nfa),
              ea = mean(ea),
              di = mean(di))
  return(summary)
}
```
## Hierarchical clustering

### loagind data
```{r}
eyes <- read.table("F:/open_git/bee_eyes/data/eyes_df.txt", header = TRUE, sep = "\t")

```
### Transforming species names into row names

#### Adding a suffix to ensure line names are unique
```{r}
row.names(eyes) <- make.unique(as.character(eyes[, 1]))
eyes <- eyes[, -1]
```
# Standardizing variables
```{r}
eyes_standardized <- scale(eyes[, 2:7])
```
## Performing hierarchical clustering
```{r}
cluster_hierarchical <- perform_hierarchical_clustering(eyes_standardized, method = "average", k = 3)
```

### Elbow and Silhouette methods for hierarchical clustering
```{r}
evaluate_clusters(eyes_standardized, hcut)
```

### Creating 3 groups of clusters
```{r}
group_eyes3 <- cutree(cluster_hierarchical, k = 3)
eyes$group_eyes3 <- group_eyes3
```
#### Saving cluster results to an Excel file

```{r}
write_xlsx(eyes, "F:/open_git/bee_eyes/results/clustereyes_average_r_.xlsx")
```
### Performing descriptive analysis by group

```{r}
mediagroup <- calculate_means_by_group(eyes, "group_eyes3")
print(mediagroup)
```
## Non-hierarchical clustering

### Running K-Means for multiple centers and plotting

```{r}
kmeans_results <- list()
for (k in 2:5) {
  result <- generate_kmeans_clusters(eyes_standardized, centers = k, title = paste("k =", k))
  kmeans_results[[paste0("k", k)]] <- result
}
```
### Arranging all K-Means cluster plots on the same screen
```{r}
grid.arrange(kmeans_results$k2$plot, kmeans_results$k3$plot,
             kmeans_results$k4$plot, kmeans_results$k5$plot, nrow = 2)
```

